---
title: "Logistic regression"
author: "Krishnan Narayanan"
date: "10 February 2017"
output: html_document
---


## Logistic regression

In this chapter we  analyse  the datasets on student alcohol consumption in secondary schools  in Portugal using Logistic regression.

The data is obtained from two questionnaires related to student alcohol consumption and is available at UCI Machine Learning Repository, [Student Alcohol consumption data page](https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION). The data consists of answers to two questionaires, one from math class and another from protugese language class related to student alcohol consumption in secondary school students in Portugal

The purpose of our analysis is to study the relationships between high/low alcohol consumption and some of the influencing  variables in the data.

## The alc dataset
Using the program create_alc.R we joined the above two datsets to a new datset alc. The alc data set has two new columns alc_use, which is the average of weekday and weekend alcohol consumption and another column high_use which is true if alc_use > 2. This dataset is also available [here](http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt)

### The variables (attributes) of alc dataframe


1. school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)
2. sex - student's sex (binary: 'F' - female or 'M' - male)
3. age - student's age (numeric: from 15 to 22)
4. address - student's home address type (binary: 'U' - urban or 'R' - rural)
5. famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)
6. Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart)
7. Medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 5th to 9th grade, 3 secondary education or 4 higher education)
8. Fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 5th to 9th grade, 3 secondary education or 4 higher education)
9. Mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
10. Fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
11. reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')
12. guardian - student's guardian (nominal: 'mother', 'father' or 'other')
13. traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14. studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
15. failures - number of past class failures (numeric: n if 1<=n<3, else 4)
16. schoolsup - extra educational support (binary: yes or no)
17. famsup - family educational support (binary: yes or no)
18. paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
19. activities - extra-curricular activities (binary: yes or no)
20. nursery - attended nursery school (binary: yes or no)
21. higher - wants to take higher education (binary: yes or no)
22. internet - Internet access at home (binary: yes or no)
23. romantic - with a romantic relationship (binary: yes or no)
24. famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25. freetime - free time after school (numeric: from 1 - very low to 5 - very high)
26. goout - going out with friends (numeric: from 1 - very low to 5 - very high)
27. Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
28. Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
29. health - current health status (numeric: from 1 - very bad to 5 - very good)
30. absences - number of school absences (numeric: from 0 to 93)
31. G1 - first period grade (numeric: from 0 to 20)
32. G2 - second period grade (numeric: from 0 to 20)
33. G3 - final grade (numeric: from 0 to 20, output target) 
34. alc_use  - average of weekday and weekend alcohol consumption
35. high_use is True if alc_use > 2


### Answer to question 2
### Reading the alc data set from amazonaws to alc dataframe

```{r echo=TRUE}
alc <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt", sep = ",", header=TRUE)
dim(alc)
summary(alc)
```


### Answer to question 2
We can see the distribution of the variables by  drawing  a bar plot using ggplot.
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyr); library(dplyr); library(ggplot2); library(GGally)
gather(alc) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()
```
### Answer to question 3
Based on the analysis we performed on the data in the datacamp exercise on this topic,  we could select the  three  attributes sex, absence, failures. As the fourth attribute I choose  final grade G3. 

### Answer to question 4



```{r echo=TRUE}
# initialise a plot of high_use and absence

g2 <- ggplot(alc, aes(x = high_use, y = absences, col = sex))
# define the plot as a boxplot and draw it
g2 + geom_boxplot() + ylab("absences") + ggtitle("Student absences by alcohol consumption and sex" )
```

From the graph of Student absences by alcohol consumption and sex, "absences" have a high 75-percentile value "high_use" group. So we select this variable.
 

```{r echo=TRUE, message=FALSE, warning=FALSE}
# initialise a plot of high_use and reason
g2 <- ggplot(alc, aes(x = high_use, y = failures, col = sex))
# define the plot as a boxplot and draw it
g2 + geom_boxplot() + ylab("failures") + ggtitle("Student failures to a select a school by alcohol consumption and sex" )
```
The above graph shows that the student failures are affected by alcohol consumption 


```{r echo=TRUE, message=FALSE, warning=FALSE}
# initialise a plot of high_use and G3
g2 <- ggplot(alc, aes(x = high_use, y = G3, col = sex))
# define the plot as a boxplot and draw it
g2 + geom_boxplot() + ylab("G3") + ggtitle("Student final grade by alcohol consumption and sex" )
```

From the Student G3(final grade) by alcohol consumption and sex graph we can observe that students with high_use of alcohol have low grades (mean and the quartile values) compared to low alcohol consumption group.

So based on the analysis described here, we have selected the variables sex, absences, failures and the final grade  G3  as the explanatory variables for the prediction on high_use 
### Answer to question 5 
Logistic regression model 

```{r echo=TRUE, message=FALSE, warning=FALSE}
m <- glm(high_use ~ absences + sex + failures + G3, data = alc, family = "binomial")

# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI <- confint(m) %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR, CI)
cbind

summary(m)
```
 Odds ratio can be used to indicate the extent to which a particular explanatory variable influences high_use.OR value close to one signifies minimal influence whereas the higher the value, the greater is the influence of the variable on the prediction outcome. So we can conclude that the explanatory variable sex has the maximum influece followed by the variables failures, absences and G3 in that order.  
 (Intercept)    absences        sexM    failures          G3 
  0.2729142   1.0956478   2.5940331   1.4133492   0.9501385 

Here we have a categorical variable sex whhich has two levels male and female.  To check whether the whole variable Sex  improves the model fit, we fit one model with (m1)  with variable sex and one without the variable sex (m2) and conduct a likelihood ratio test. This tests the hypothesis that all coefficients of sex are zero:
```{r echo=TRUE, message=FALSE, warning=FALSE}
m1 <- glm(high_use ~ absences + sex + failures + G3, data = alc, family = "binomial")
m2 <- glm(high_use ~ absences + failures + G3, data = alc, family = "binomial")

anova(m1, m2, test="LRT")
```

The likelihood ratio test is highly significant and we would conclude that the variable sex should remain in the model.

### Answer to question 6
Logistic regression enables to predict binary outcomes. We use the data alc to make the predictions for high:use. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
m <- glm(high_use ~ absences + sex + failures + G3, data = alc, family = "binomial")
# predict() the probability of high_use
probabilities <- predict(m, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = probability > 0.5)
# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction)
```
We see from the results that the high_use prediction is fairly accurate when the high_use has the state FALSE, and the prediction is rather poor when high_use has the state TRUE. 
